% CREATED BY DAVID FRISK, 2016
\chapter{Conclusion}

This thesis investigated the problem of compressing multimodal automotive and IoT time-series data while preserving information relevant for downstream machine learning tasks. Through systematic evaluation of discrete and continuous neural compression methods across two datasets and two downstream applications, this work provides empirical insights into the fundamental trade-offs between compression efficiency and task utility in edge computing settings.

\section{Main Findings}

Discrete-latent neural compression methods achieve substantial compression ratios (200x+) on multimodal time-series data while maintaining acceptable reconstruction quality. However, this work reveals that multimodal data composition creates significant challenges: different signal types with contrasting characteristics are difficult to handle uniformly, and no single compression strategy excels equally across all signal types. Continuous-latent baselines, such as TCN-RNN, fundamentally cannot compete at similar compression ratios, demonstrating the practical superiority of discrete bottlenecks for edge compression.

Specializing compression architectures for multimodal data, whether through partitioned codebooks or task-aware training, gives insights into how shaping the optimization landscape impacts the performance of neural compression models. Constraining the model by partitioning the codebook can improve consistency and reduce variance across random seeds but may sacrifice flexibility when signals have varying importance across different downstream tasks. Task-aware training occasionally achieves strong utility retention but introduces training instability and reproducibility challenges. 

Utility retention proves to be highly variable and task-dependent. While superior reconstruction correlation does not guarantee better downstream utility, the empirical results demonstrate that signal-specific characteristics matter significantly. In the anomaly detection task, reconstruction fidelity correlates strongly with utility, whereas in regression tasks on noisy data, task-specific signal importance becomes more critical. This finding underscores that modern neural compression methods that produce data for downstream models must be evaluated through both reconstruction and downstream metrics, as optimization for one does not automatically optimize the other.

\section{Key Contributions} 

This thesis makes three primary contributions. First, the thesis provides, through both the TAEdgeCodec and the UniEdgeCodec architectures, empirical evidence for the nuanced complexitiy of multimodal neural compression and demonstrates the fundemental trade-offs that need to be accounted for. Second, empirical evidence for the limitations of continuous-latent neural compression techniques for edge deployment on multimodal data is presented. Third, it establishes that compression method selection must be informed by downstream application requirements, and that reconstruction metrics alone are insufficient for method evaluation. 

\section{Future work}
The results of this thesis highlight the importance of considering both data characteristics, model architecture and optimization objectives when designing neural compression methods for multimodal IoT data. Although the results suggest that constrained models, such as the UniEdgeCodec, can improve upon stability while limiting the learning capacity, the results are not conclusive. Future work focusing on ablation studies of the UniEdgeCodec architecture have potential to shed light on the trade-offs between stability and learning capacity. Another possible avenue for future work is to explore variations of the TAEdgeCodec and UniEdgeCodec architectures that attempt to mitigate the stability issues observed in the TAEdgeCodec model, while still allowing for flexibility in terms of learning goals. Combining the partitioned codebook design of the UniEdgeCodec with a task loss term, for example, could be a promising direction to explore. 

As this evaluation was limited to laboratory settings with only two downstream tasks (regression and anomaly detection) and since it did not include real hardware deployment or computational efficiency measurements, future work should also verify the efficacy of the presented methods in a real world setting. Lastly, future work could explore adaptive signal classification using learned approaches. The manual signal classification is a clear limitation of the presented UniEdgeCodec method and exploring the different signal types more closely should.
