% CREATED BY DAVID FRISK, 2016
\chapter{Discussion}

<<<<<<< HEAD
This chapter provides further discussion and interpretation of the results previously presented in the Results chapter. 

\subsection{Research Questions} 

The results can be synthesized with respect to the three research questions. RQ1 relates to the performance of the selected neural compression methods on multimodal IoT data. While this question is broad and difficult to answer in a fully general sense, the results show that the discrete methods, EdgeCodec, UniEdgeCodec and TAEdgeCodec, are able to achieve high CRs of 200x+ while maintaining good reconstruction quality. The results also highlight that multimodal data creates additional challenges for compression methods, since the signals can have very different characteristics and therefore require tailored compression strategies. 

RQ2 relates to the utility retention of the selected neural compression methods. As can be seen in the results, the utility retention is highly variable across random seeds, and the relationship between reconstruction quality and utility retention is complex and task-dependent. It is hard to definitively say which architecture favours utility retention the most, but it is clear that minimizing the loss of task relevant information is difficult at high CRs. 

RQ3 relates to the limitations of the existing methods (EdgeCodec) and how to mitigate these limitations. The results here show that the partitioned codebook design of the UniEdgeCodec is somewhat effective at improving signal reconstruction quality, in terms of correlation, training stability and perceptual quality but does not always lead to improved utility retention. Using a task loss term, as in the TAEdgeCodec, can lead to improved utility retention, but struggles with training stability and reconstruction quality generally suffers as a result. 

\subsection{Interpretation of Empirical Results} 
=======
\textbf{Research Question Synthesis} The results can be synthesized with respect to the three research questions. RQ1 relates to the performance of the selected neural compression methods on multimodal IoT data. While this question is broad and difficult to answer in a fully general sense, the results show that the discrete methods, EdgeCodec, UniEdgeCodec and TAEdgeCodec, are able to achieve high CRs of 200x+ while maintaining good reconstruction quality. The results also highlight that multimodal data creates additional challenges for compression methods, since the signals can have very different characteristics and therefore require tailored compression strategies. 
>>>>>>> d3c6ef03dc18805843519cd8bfab47e0cd776752

RQ2 relates to the utility retention of the selected neural compression methods. As can be seen in the results, the utility retention is highly variable across random seeds, and the relationship between reconstruction quality and utility retention is complex and task-dependent. It is hard to definitively say which architecture favours utility retention the most, but it is clear that minimizing the loss of task relevant information is difficult at high CRs. 

RQ3 relates to the limitations of the existing methods (EdgeCodec) and how to mitigate these limitations. The results here show that the partitioned codebook design of the UniEdgeCodec is somewhat effective at improving signal reconstruction quality, in terms of correlation, training stability and perceptual quality but does not always lead to improved utility retention. Using a task loss term, as in the TAEdgeCodec, can lead to improved utility retention, but struggles with training stability and reconstruction quality generally suffers as a result. 

\textbf{Interpretation of Empirical Evidence} The superior reconstruction correlation achieved by UniEdgeCodec (0.960 $\pm$ 0.007 on EMOB) compared with its competitors can be explained by its design objective of optimizing perceptual quality, particularly for irregular and discontinuous signals. Because perceptual quality is closely correlated with reconstruction correlation, as can be seen in the visualizations of the reconstructed signals, explicit optimization for this objective should lead to improved correlation metrics across all signal types. This consistent preservation of temporal structure is especially beneficial for downstream anomaly detection tasks, where classification reliability depends on accurately capturing temporal patterns. 

The model's architectural choice of partitioning the codebook, and thus preventing signals from competing within the same codebook, together with the regularizing effect of the loss function, contributes to the training stability observed across all random seeds. These factors likely resulted in the UniEdgeCodec having both the lowest metric variance and the most reliable convergence behavior across the experimental runs. 

This being said, there are notable downsides to having a partitioned codebook, as is exemplified in both use case. UniEdgeCodec consistently achieves the lowest utility retention in the appliances energy use case, mainly, because of the poor reconstruction of the outside temperature. When calculating the pearson correlation between the ``T\_out'' signal, and utility retention for the appliances energy use case (r=0.814, p=0.0076), it becomes clear that the outside temperature is massively important for this use case. A similar behavior can be seen in the Volvo EMOB use case. Here the overall reconstruction quality is even better for the UniEdgeCodec compared to the TAEdgeCodec and the EdgeCodec models. But the utility retention is still better, although unreliably so, for the TAEdgeCodec and EdgeCodec models. Again, individual signals seem to be unevenly important for the downstream task. While EdgeCodec in most cases reconstructs ``Temp 1'' adequately, this comes at the cost of poor reconstruction of channels with sparse, irregular patterns like the ``Amps'' signal. Looking at the seed where EdgeCodec performs worse in terms of utility retention for the Volvo EMOB use case, it can be seen that the correlation for the ``Temp 1'' signal degrades to 0.93, compared with a correlation of 0.99 for seed 42, while the correlation for the ``Amps'' signal improves to 0.34, compared with a correlation of 0.32 for seed 42. This points to a correlation between ``Temp 1'' and the anomalies to be detected, but it also shows that the EdgeCodec architecture fundamentally has to trade off significantly between reconstructing smooth, continuous signals and irregular, discontinuous signals. For the UniEdgeCodec model, this means that even though it can achieve a higher total correlation and comparable or better reconstruction MSE on many signals, specific signals matter more for certain downstream tasks than others.

The mixed and seed-dependent results of TAEdgeCodec would suggest that multi-objective neural compression struggles when with producing stable result. This reveals a fundamental challenges of models such as this one. The more complex optimization landscape that arises from jointly learning compression and downstream task performance creates inherent trade-offs. The model must allocate limited representational capacity between reconstruction fidelity and task-specific features. This likely explains the observed pattern of worse reconstruction compared with UniEdgeCodec and EdgeCodec, instability in utility retention, and occasional convergence failures, such as the early collapse observed in seed 100 for the Volvo EMOB data. That said, the task-aware nature of TAEdgeCodec does occasionally lead to very good utility retention results. This suggests that, when the optimization objectives can be jointly satisfied, the approach can succeed. However, the substantially higher variance across seeds indicates that the balance between reconstruction and task optimization is fundamentally unstable without the additional architectural constraints provided by UniEdgeCodec's per-signal codebook design. It is fair to say that, for this model, multimodal data compression and task-aware learning exceed the model's representational capacity.

EdgeCodec's strong reconstruction performance in terms of mean MSE across both datasets (0.038 $\pm$ 0.028 on EMOB, 0.490 $\pm$ 0.172 on appliances energy) is consistent with its primary optimization objective of reconstruction fidelity. However, its utility retention results reveal, that it is highly dependent on focusing on the relevant signals.

The TCN-RNN baseline demonstrates that continuous latent representations become increasingly difficult at high CRs. Operating at only 3.0x compression with a correlation of 0.726 on EMOB, which is substantially below the 200x+ compression of discrete methods, it shows that continuous bottleneck compression is inherently limited for sequence compression tasks. To be competitive, the architecture of the TCN-RNN baseline would need to be significantly more complex, making the model heavier. While discrete latent representations provide a clear advantage, the baseline results also underscore that model objectives matter significantly. The superior performance of the discrete methods stems not only from discrete latents, but also from the specific optimization objectives designed for each architecture.

The relationship between reconstruction quality and downstream utility retention exhibits important insights into the validity of the different metrics used to measure the performance of the neural compression methods. For the EMOB anomaly detection task, reconstruction MSE serves as a reliable predictor of utility retention. This might be, because the task fundamentally depends on capturing sparse, discontinuous events in the signal. When these critical events are not reliably reconstructed, as evidenced by poor MSE, downstream anomaly detection inherently suffers. In contrast, the appliances energy data has more signals and therefore also more noise in the downstream task. Not all of these signals are likely to correlate strongly with the regression target, making the overall reconstruction MSE less meaningful as a predictor for donwstream task utility retention, and signal-specific metrics more relevant. This task-dependent behavior highlights that compression suitability cannot be assessed universally through reconstruction metrics alone. It is vital that the choice of compression method be informed by the specific requirements of the downstream application, and that compression methods are specifically evaluated using both reconstruction quality and utility retention metrics.

<<<<<<< HEAD
\subsection{Methodological Implications}

Two central implications emerge from the experimental results and the preceding analysis. First, continuous latent autoencoders, at least those with deterministic encoders, do not have the capacity to achieve satisfactory multimodal compression results while remaining lightweight enough to be edge-compatible. The TCN-RNN model is fundamentally constrained in its capabilities to achieve competitive CRs, and even at much lower CRs it cannot compete with discrete-latent neural compression models in terms of reconstruction quality. 
=======
\textbf{Methodological Implications} Two central implications emerge from the experimental results and the preceding analysis. First, continuous latent autoencoders, at least those with deterministic encoders, do not have the capacity to achieve satisfactory multimodal compression results while remaining lightweight enough to be edge-compatible. The TCN-RNN model is fundamentally constrained in its capabilities to achieve competitive CRs, and even at much lower CRs it cannot compete with discrete-latent neural compression models in terms of reconstruction quality. 
>>>>>>> d3c6ef03dc18805843519cd8bfab47e0cd776752

The second important takeaway is that, when constructing neural compression methods, it is vital to be conscious of the optimization landscape. The experiments make clear that the limits of representational capacity for edge-optimized neural compression methods are real and must be accounted for. When constraining the optimization landscape, for example through the introduction of a partitioned codebook, stability can be gained. This is indicated by the stability observed in the UniEdgeCodec model. But importantly, this stability comes with a trade-off. When loosening the optimization landscape, for example by adding a task loss term or a shared codebook, models become less reliable but also less constrained. The higher peak utility retention of both TAEdgeCodec and EdgeCodec and the signal specific superior reconstruction performance shows that these models benefit from this freedom, whereas UniEdgeCodec is more constrained. The freedom to be able to not only learn correlating signal patterns in a shared optimization landscape, but also assign these correlating patterns to a shared codebook, can improve the reconstruction of correlated patterns. For neural compression methods that produce data for downstream model training, this means that being conscious of how the downstream task goals and the training goals align is vital. It is important to one be aware of the importance of specific signals for the downstream tasks, and two, take the trade-off between stability and model freedom into account. It has further been shown, that a holistic evaluation including both reconstruction metrics and utility retention metrics is need to fully understand the efficacy of neural compression methods that provide compressed data for downstream model training.