% CREATED BY DAVID FRISK, 2016
\chapter{Introduction}

\section{Aim}
The aim of this project will be to investigate the use of neural compression techniques, specifically tokenization-based approaches, for compressing automotive time series data. The goal is to develop a compression framework that effectively balances the trade-off between compression ratio and utility for downstream machine learning (ML) tasks, such as predictive maintenance and anomaly detection.

The motivation behind using tokenization as a neural compression strategy for automotive time series data is to produce discrete latent representations that simplify the entropy modeling task within the compression pipeline. By constraining the data to a finite set of tokens, the complexity of modeling the underlying data distribution is reduced, enabling the use of lightweight entropy models that are computationally efficient. This is particularly advantageous in automotive applications where computational resources are limited, and real-time processing is often required.

The goal will be to design, implement and evaluate a tokenization-based neural compression framework tailored for time-series data. Evaluation will be conducted by comparing relevant parameters between the proposed framework and traditional compression methods. 

\section{Problem}
The problem to be solved with this thesis is that of efficiently compressing vehicle telemetry data for downstream machine learning tasks. The issue is that vehicles generate a vast amount of telemetry data from various sensors and systems, which can be very challenging to store and transmit due to constraints on edge devices and communication bandwidth \parencite{samantaray2023}. While techniques for data compression do exist, they are either not optimized for the specific characteristics of vehicle telemetry data or do not take into account the requirements of downstream machine learning tasks.

Traditional compression techniques such as event-triggered logging and algorithmic time series compression methods have limitations when applied to vehicle telemetry data. Event-triggered logging can miss important information that does not cross predefined thresholds, leading to incomplete data for machine learning models \parencite{nunes2023, jimenez2020, azar2022}. Algorithmic compression methods often rely on manually chosen parameters and may not effectively capture the complex patterns present in vehicle telemetry data, which can negatively impact the performance of downstream machine learning tasks. 

More recent approaches such as utility-aware adaptive telemetry and task-aware compression have shown promise in addressing some of these challenges. However, these methods often require complex policy learning or adaptive algorithms that may not be feasible for real-time applications in vehicles \parencite{zheng2023, löhdefink2019, kawawabeaudan2022, liu2024deepdictdeeplearningbased}.

One less explored but promising approach to compressing data which could be suitable for vehicle telemetry data is that of neural compression in combination with tokenization. A system which incorporates a lightweight neural encoder and a small tokenizer should in theory be able to produce a lossy low entropy representation of the data which can be efficiently stored and transmitted. The main challenge will be to ensure that the compressed representation retains as much relevant information as possible for downstream machine learning tasks while still maintaining a high compression ratio.

\section{Limitations}
Due to time constraints and the scope of this thesis, there will be several limitations to the work presented:
\begin{itemize}
    \item The work will be limited to only testing a single head / type of loss for the neural compression model. While it would be interesting to explore multiple heads, this would be time consuming in testing when running all the model variations. 
    \item The work will only implement the encoder and tokenizer part of the pipeline while forgoing the entropy coding. This is done due to the focus of the thesis being on the neural compression step and while entropy coding is important in a full compression pipeline, it is less relevant for the research question at hand. 
    \item The work will also forgo testing on real hardware as that would be very time consuming and difficult to set up while contributing relatively little to the main research question.
    \item The work will be limited to only testing regression tasks and anomaly detection as downstream tasks due to time constraints and structure of the available datasets. These two tasks should however be a good representation of common downstream tasks and adding more tasks would likely not drastically change the conclusion of the thesis. 
    \item Due to wanting to keep the compressed data representation as versatile as possible, no task-aware compression techniques will be implemented. While task-aware compression could potentially yield better performance for specific tasks, it would limit the generalizability of the compressed representation and add significant complexity to the work. It also results in a more relevant comparison to other compression techniques which are not task-aware. 
\end{itemize}

% This chapter presents the section levels that can be used in the template. 

% \section{Section levels}
% \autoref{tab:sections} presents an overview of the section levels that are used in this document. The number of levels that are numbered and included in the table of contents is set in the settings file \texttt{Settings.tex}. The levels are shown in Section \ref{Section_ref}.

% This is a new paragraph and should have proper parskip or indentation. Don't forget to cite your sources~\cite{Brajnik2008}. % '~' becomes space which cannot line break.

% \begin{table}[h]
% \centering
% \caption{Section levels} % Table text above table.
% \begin{tabular}{ll} \hline
% Name & Command\\ \hline
% Chapter & \textbackslash\texttt{chapter\{\emph{Chapter name}\}}\\
% Section & \textbackslash\texttt{section\{\emph{Section name}\}}\\
% Subsection & \textbackslash\texttt{subsection\{\emph{Subsection name}\}}\\
% Subsubsection & \textbackslash\texttt{subsubsection\{\emph{Subsubsection name}\}}\\
% %Paragraph & \textbackslash\texttt{paragraph\{\emph{Paragraph name}\}}\\
% %Subparagraph & \textbackslash\texttt{paragraph\{\emph{Subparagraph name}\}}\\ \hline\hline
% \end{tabular}
% \label{tab:sections}
% \end{table}

% \section{Section} \label{Section_ref}
% \subsection{Subsection}
% \subsubsection{Subsubsection}
% %\paragraph{Paragraph}
% %\subparagraph{Subparagraph}

